Covariate analysis of ROSMAP Reprocessed RNASeq (with SVA)

### 2016-05-31 code/Rmd/MSSM_RNASeq_SVA_cdr.Rmd
Added two-step SVA based processing for MSSM RNASeq (all four brain regions)

---
title: "Covariates analysis of MSBB RNASeq data"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)
library(githubr)

@@

knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_cdr.Rmd",
                                 parentId ="syn5570248",
                                 entityName = 'MSBB RNASeq (based on CDR)')
```

@@

thisFileName <- 'MSSM_RNASeq_cdr.Rmd'

@@

# Select covariates of interest
FactorCovariates <- c('RACE', 'SEX', 'BrainRegion.CDR', 'Batch', 'BB')
ContCovariates <- c('PMI', 'AOD', 'RIN', 'RIN2', 'PlaqueMean')

@@

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix. Here BB (Donor Information) is chosen as random effect
```{r iterativeNorm}
# Primary variable of interest
postAdjustCovars = 'BrainRegion.CDR'

@@

covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, 'BB'))]

postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))

@@

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`
### Sanity check
```{r residualSigCovars.manual, fig.width=20, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Residual calculation
Calculate weighted residuals and add back "BrainRegion.CDR" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("BrainRegion.CDR", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```

@@

tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
cols = WGCNA::labels2colors(COVARIATES.tmp);
tmp = gsub('\\..*','',cols)
WGCNA::plotDendroAndColors(tree, 
                           colors = tmp, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```
```{r get.gene.symbols, cache=TRUE, eval=FALSE}
# Define biomart object
mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl", host = )

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = rownames(RESIDUAL.GENE_EXPRESSION),
                       mart = mart)
RESIDUAL.GENE_EXPRESSION <- left_join(rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, "ensembl_gene_id"),
                                      Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
VOOM.ADJUSTED.GENE_EXPRESSION$E <- left_join(rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E, "ensembl_gene_id"),
                                             Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = as.data.frame(summary(COVARIATES$BrainRegion.CDR))
colnames(tmp) = 'No. of samples'
kable(tmp)
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=18}
# Fit contrast
contrast = makeContrasts(contrasts = sapply(unique(METADATA$BrainRegion), function(x){
  apply(combn(paste0('BrainRegion.CDR',x,'.',unique(METADATA$CDR)),2), 2, paste, collapse = '-')
}) %>%
  as.character(), 
levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
FIT = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
FIT = eBayes(FIT)

# Get differnetial expression
DE = list()
p = list()
for (i in 1:dim(contrast)[2]){
  DE[[i]] = topTable(FIT, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
  DE[[i]] = rownameToFirstColumn(DE[[i]], "ensembl_gene_id")
  p[[i]] <- ggplot(DE[[i]], aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
  p[[i]] <- p[[i]] + ggtitle(gsub('BrainRegion.CDR','',colnames(contrast)[i])) + xlim(c(-1.5,1.5)) + ylim(c(0,6))
  p[[i]] <- p[[i]] + geom_hline(yintercept = -log10(0.05), color = 'red')
}
# multiplot(plotlist=p, cols = 4)
names(DE) = gsub('BrainRegion.CDR','',colnames(contrast))

DE = DE %>% ldply(.id = 'Comparison')

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(Comparison, ensembl_gene_id) %>%
  group_by(Comparison) %>%
  dplyr::summarise(nDiffExp = length(unique(ensembl_gene_id)),
                   DiffExpGenes = paste(unique(ensembl_gene_id), collapse = ','))

kable(tmp[,1:2])
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "MSBB RNASeq (based on CDR)", parentId = parentId)
CODE <- synStore(CODE)

>>>>

---
title: "Covariates analysis of MSBB RNASeq (with SVA)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)
library(githubr)

@@

knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_SVA_cdr.Rmd",
                                 parentId ="syn5570248",
                                 entityName = 'MSBB RNASeq (based on CDR with SVA)')
```

@@

thisFileName <- 'MSSM_RNASeq_SVA_cdr.Rmd'

@@

# Select covariates of interest
FactorCovariates <- c('RACE', 'SEX', 'BrainRegion.CDR', 'Batch', 'BB')
ContCovariates <- c('PMI', 'AOD', 'RIN', 'RIN2')

@@

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix. Here BB (Donor Information) is chosen as random effect
```{r iterativeNorm}
# Primary variable of interest
postAdjustCovars = c('BrainRegion.CDR', 'RIN', 'PMI', 'AOD', 'Batch') # Based on a separate analysis

@@

# covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, 'BB'))]
# 
# postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))

@@

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`
### Sanity check
```{r residualSigCovars.manual, fig.width=20, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Estimate surrogate variables using SVA package
```{r sva}
MODEL0 = DM1$design[,-grep('BrainRegion.CDR',colnames(DM1$design))] # Get null model by removing variable of interest
MODEL1 = DM1$design

## Calculate the number of surrogate variables
NUM.SV = num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, method = 'be', seed = 123456)

SURR.VAR = sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0, method = "two-step", n.sv = NUM.SV, B = 30)$sv
colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])

# Estimate correlation of random effects
VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=cbind(MODEL1, SURR.VAR), plot=F)
correlation <- parallelDuplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION, block=COVARIATES$BB)

# Re-calculate voom weights with correlation of random effects
VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=cbind(MODEL1, SURR.VAR), plot=F,
                                     block=COVARIATES$BB, correlation = correlation$cor)

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
```

### Residual calculation
Calculate weighted residuals and add back "BrainRegion.CDR" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("BrainRegion.CDR", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```

@@

tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
cols = WGCNA::labels2colors(COVARIATES.tmp);
tmp = gsub('\\..*','',cols)
WGCNA::plotDendroAndColors(tree, 
                           colors = tmp, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```
```{r temp1, include=FALSE}
dev.off()
gc()
```

```{r get.gene.symbols, cache=TRUE, eval=FALSE}
# Define biomart object
mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl", host = )

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = rownames(RESIDUAL.GENE_EXPRESSION),
                       mart = mart)
RESIDUAL.GENE_EXPRESSION <- left_join(rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, "ensembl_gene_id"),
                                      Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
VOOM.ADJUSTED.GENE_EXPRESSION$E <- left_join(rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E, "ensembl_gene_id"),
                                             Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = as.data.frame(summary(COVARIATES$BrainRegion.CDR))
colnames(tmp) = 'No. of samples'
kable(tmp)
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=25}
# Fit contrast
contrast = makeContrasts(contrasts = sapply(unique(METADATA$BrainRegion), function(x){
  apply(combn(paste0('BrainRegion.CDR',x,'.',unique(METADATA$CDR)),2), 2, paste, collapse = '-')
}) %>%
  as.character(), 
levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
FIT = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
FIT = eBayes(FIT)

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC){
  topTable(FIT, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1]) %>%
    rownameToFirstColumn("ensembl_gene_id") %>%
    left_join(Ensemble2HGNC)
}, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC)
names(DE) = gsub('BrainRegion.CDR','',colnames(contrast))

DE = DE %>% 
  ldply(.id = 'Comparison') %>%
  dplyr::mutate(Comparison1 = Comparison) %>%
  tidyr::separate(Comparison1, c('from.state','to.state'), sep = '-') %>%
  tidyr::separate(from.state, c('BrainRegion','reference'), sep = '\\.') %>%
  tidyr::separate(to.state, c('BrainRegion','against'), sep = '\\.') %>%
  tidyr::unite(Name, reference, against, sep = '-')

p <- ggplot(DE, aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
p <- p + geom_hline(yintercept = -log10(0.05), color = 'red')
p <- p + geom_vline(xintercept = log2(1.2), color = 'red')
p <- p + geom_vline(xintercept = -log2(1.2), color = 'red')
p <- p + facet_grid(BrainRegion~.+Name)
p

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(BrainRegion, Name, ensembl_gene_id) %>%
  group_by(BrainRegion, Name) %>%
  dplyr::summarise(nDiffExp = length(unique(ensembl_gene_id))) %>%
  tidyr::spread(Name, nDiffExp)

kable(tmp)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "MSBB RNASeq (based on CDR with SVA)", parentId = parentId)
CODE <- synStore(CODE)

### 2016-05-31 code/Rmd/ROSMAP_RNASeq_SVA_Reprocessed.Rmd
Added two-step SVA based processing for MSSM RNASeq (all four brain regions)

knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_SVA_Reprocessed.Rmd",
                                 parentId ="syn5570291",
                                 entityName = 'ROSMAP RNASeq Reprocessed Counts (Based on cogdx with SVA)')
```

@@

SURR.VAR = sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0, method = "two-step", n.sv = NUM.SV, B = 30)$sv
colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])

@@

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC)) + geom_point() + xlim(c(-1,1))
p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
p = p + geom_vline(xintercept = -log2(1.2), color = 'red') + facet_grid(.~Comparison, scales = 'fixed')
p
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "ROSMAP RNASeq Reprocessed Counts (Based on cogdx with SVA)", parentId = parentId)
CODE <- synStore(CODE)

>>>>

knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_SVA_Reprocessed.Rmd",
                                 parentId ="syn5570291",
                                 entityName = 'ROSMAP RNASeq Reprocessed Counts (Based on cogdx with SVA IRW)')
```

@@

SURR.VAR = sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0, method = "irw", n.sv = NUM.SV, B = 30)$sv
colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])

@@

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC)) + geom_point() + xlim(c(-1,1))
p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
p = p + geom_vline(xintercept = -log2(1.2), color = 'red') + facet_grid(.~Comparison, scales = 'fixed')
p
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "ROSMAP RNASeq Reprocessed Counts (Based on cogdx with SVA IRW)", parentId = parentId)
CODE <- synStore(CODE)
